---
title: "Progression of joint destruction in rheumatoid arthritis"
author: "Group 2"
date: "2025-03-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,error = FALSE, warning = FALSE)
```

Package library
```{r }
library(nlme) 
library(lme4)
library(lattice)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(lme4)
library(lattice)
library(car)
library(GLMMadaptive)
library(gridExtra) 
library(patchwork)
```

# Basic information about the dataset
```{r }
load("data2.RData")  # Data import. The dataset is in long format. No need to transform.
data = DF              
str(data)   #  data structure
dim(data)   #  2723    8
# range(data$SHS) # 0,119, Sharp index at each of the 8 visits, respectively.
unique(data$Severity) # Disease severity (0: low, 1: high) at each of the 8 visits, respectively.
unique(data$Visit)  # 0,1,2,3,4,5,6,7 visits by year
unique(data$Period) # inclusion period,  period A:before "1996-01-01"; period B:  after "1996-01-01" 
range(data$Age) # 41.2~87.9, age of the patient at inclusion.
unique(data$sex) #  1: males, 2: females
unique(data$SNP) # single nucleotide polymorphism (0: no variant alleles, 1: variant allele, 2: variant alleles).
```
## 1.Missing data
It is introduced that "Not all planned measurements have been recorded for several reasons e.g., loss to follow-up, relocation, remission, etc". In dataset, there is no missing data in observed dataset. We deem it as missing-at-random (MAR) and missing completely-at-random (MCAR) conditions. 
Missing data distribution table: 
```{r }
missing_rows <- sum(rowSums(is.na(data)) > 0)
missing_rows # 0. There is no missing data in observed dataset. 
cat("\n### Visit Distribution\n")
table(data$Visit)  # Count of observations per visit time point
prop.table(table(data$Visit))  # Proportion per visit
```

Missing data distribution graphing: 
```{r }
visit_counts <- data %>%
  group_by(Visit) %>%
  summarise(N = n())

ggplot(data.frame(Visit = c(0:7), Count = c(500, 326, 334, 307, 327, 321, 299, 309)), 
       aes(x = Visit, y = Count)) +
  geom_line() +
  geom_point() +
  geom_text(aes(label = Count), vjust = -0.5, size = 5) + 
  labs(title = "Number of Patients Over Time", x = "Visit", y = "Number of Patients")+
  ylim(c(200, 550))
```
After the baseline visit, different combinations of patients, each consisting of about 63.5% ± 2.5% of the base-line visited patients, has participated in each visit, thereafter.
It is introduced that "Not all planned measurements have been recorded for several reasons e.g., loss to follow-up, relocation, remission, etc". 
These missingness reasons have nothing to do with the missed outcomes, which suggests that perhaps the mechanism is not MNAR. Exploring the data shows there are both dropped-outs and intermittents in the data: 472 Intermittent Missingness and 28 Dropouts.
But why there is no NA in the data? because firstly this is a Long dataset, not a Wide one! NA's will apear if we convert it from long to wide, and secondly, when a patient has missed a visit, the value of no variable (not even his id) is recorded for him on that date into the dataset, thus no NA for him on that date!
but if we use likelihood-based methods to estimate model parameters, then the missingness can be ignored, provided that the missingness mechanism is MCAR or MAR. So, for subject-specific inferences we can use GLMM, but for overall inferences we cannot use GEE (because GEE is not a likelihood approach), but we have to derive these inferences from GLMM results or by GLMMadaptive approximation.


## 2.Transform variables into factor
```{r }
data$Severity.f = factor(data$Severity)
data$SNP.f = factor(DF$SNP)
data$log_SHS <- log(data$SHS + 1)
# head(data)  # keep visit continuous variable. 
```

# Sharp index Analysis
## 1.Descriptive Analysis
### 1.1 mean structure
```{r}
 mean_SHS <- data %>% group_by(Period, Visit) %>%
  summarise(mean_SHS = mean(SHS, na.rm = TRUE),sd_sharp = sd(SHS, na.rm = TRUE),n = n(),.groups = "drop")%>% mutate(se = sd_sharp / sqrt(n))
mean_SHS_SNP <- data %>% group_by(Period, SNP,Visit) %>% summarise(mean_SHS = mean(SHS, na.rm = TRUE),n = n(),.groups = "drop")
# Mean SHS by Period
p1 <-ggplot(mean_SHS, aes(x = Visit, y = mean_SHS, color = Period, group = Period)) +
  geom_line(linewidth = 1.2) +geom_point(size = 2) + labs(title = "Mean SHS Over Time by Period",x = "Visits",y = "Mean SHS") +
  geom_errorbar(aes(ymin = mean_SHS - se, ymax = mean_SHS + se), width = 0.2) +
  labs(title = "", y = "Mean Sharp Index", x = "Visits") +theme_minimal()
# Mean SHS Over Period by SNP
p2 <-ggplot(mean_SHS_SNP, aes(x = Visit, y = mean_SHS, color = Period, group = Period)) +
  geom_line(size = 1.2) +geom_point(size = 2) +facet_wrap(~ SNP) +
  labs(title = "",x = "Visits",y = "Mean Sharp Index (SHS)") +
  theme_minimal() +theme(legend.position = "top")  
# p1 + p2 
```

### 1.2 random effect
```{r }
IDs_A <- unique(data$id[data$Period == "A"])   # Individual changes in SHS by period.
IDs_B <- unique(data$id[data$Period == "B"])
set.seed(123)  # reproducibility   
sample_A <- sample(IDs_A, 10)  # How many to sample from each period? (Here, 10 from A and 10 from B)
sample_B <- sample(IDs_B, 10)
sampled_IDs <- c(sample_A, sample_B)  # Combine sampled IDs
DF_sampled <- subset(data, id %in% sampled_IDs)  # Filter dataset
p1 <-xyplot(SHS ~ Visit | Period,groups = id,data = DF_sampled,type = "l",xlab = "Visit",ylab = "SHS",auto.key = list(title = "id", space = "up"),panel = function(x, y, groups, ...) {panel.superpose(x, y, groups, ...)})
IDs_E <- unique(data$id[data$SNP == 0])     # Individual changes in SHS by SNP
IDs_F <- unique(data$id[data$SNP == 1])
IDs_G <- unique(data$id[data$SNP == 2])
sample_E <- sample(IDs_E, 10) # How many to sample from each period? (Here, 10 from A and 10 from B)
sample_F <- sample(IDs_F, 10)
sample_G <- sample(IDs_G, 10)
sampled_IDs <- c(sample_E, sample_F, sample_G)  # Combine sampled IDs
DF_sampled <- subset(data, id %in% sampled_IDs) # Filter dataset
p2 <-xyplot(SHS ~ Visit | SNP,groups = id,data = DF_sampled,type = "l",xlab = "Visit",ylab = "SHS", auto.key = FALSE,
            panel = function(x, y, groups, ...) {panel.superpose(x, y, groups, ...) })  # Create the plot
# grid.arrange(p1, p2, ncol = 2)
```

## 2.Model Building and selection
### 2.1 Initial model 
```{r}
# fit an elaborate model according to EDA, we try to capture non-linear trend of the data and we fit quadratic model.
model1 <- lme(SHS ~ (Period + Age + sex + SNP.f) * (I(Visit) + I(Visit^2)), data = data, random = ~ 1| id,method = "REML")
p1 <-plot(model1, resid(., type = "n") ~ fitted(.), type = c("p", "smooth"),col = "#A1C9F4",lwd = 3)  # check Residuals vs Fitted values
p2 <-qqnorm(model1, ~ resid(., type = "n"), abline = c(0, 1),col = "#A1C9F4")   # check  qq plot
# grid.arrange(p1, p2, ncol = 2)
```
### 2.2 Hypothesis testing for covariances
```{r}
# Fit LMM with random intercepts and random slopes
model_slope <- lme(log_SHS ~ (Period + Age + sex + SNP.f) * Visit, data = data, random = ~ Visit| id,method = "REML")  
# Fit LMM with only random intercepts
model_intercept <- lme(log_SHS ~ (Period + Age + sex + SNP.f) * Visit, data = data, random = ~ 1| id,method = "REML")  
# Compare the two models with LRT. The result is significant, which means model_slope is better.
lrt <- anova(model_slope, model_intercept)
# 0.5*pchisq(lrt$ L.Ratio[2], df=1, lower.tail = FALSE) + 0.5*pchisq(lrt$ L.Ratio[2], df=2, lower.tail = FALSE)
```
### 2.3 Hypothesis testing for fixed effects
```{r}
# extract t_table of estimated parameters  # summary(model_slope)$tTable
# Fit a simplified mean model with random intercept and slope, and this our final model.
model_new <- lme(log_SHS ~ Period + sex + Visit + sex * Visit, data = data, random = ~ Visit| id,method = "REML")   
```
### 2.4 Estimated variances and covariance of randoms effects
```{r}
vcov_mat <- getVarCov(model_new)[1:2, 1:2]
total_variance <- vcov_mat[1,1] + vcov_mat[2,2] + 2 * vcov_mat[1,2]
```
### 2.5 Model diagnostics
```{r}
p3 <-plot(model_new, resid(., type = "n") ~ fitted(.),type = c("p", "smooth"),col = "#A1C9F4",lwd = 3) # Residuals vs Fitted values
# Residuals vs Fitted values per gender group
p4 <-plot(model_new, resid(., type = "n") ~ fitted(.) | sex, type = c("p", "smooth"), col = "#A1C9F4",lwd = 3) 
# Residuals vs Fitted values per period
p5 <-plot(model_new, resid(., type = "n") ~ fitted(.) | Period, type = c("p", "smooth"), col = "#A1C9F4",lwd = 3)
# grid.arrange(p3, p4, p5,ncol = 3)
```

```{r}
residual_con <- residuals(model_new,type = "n",level = 1)
# Extract residuals and Visit variable
data1 <- data.frame(id = data$id,Visit = data$Visit,Period = data$Period,sex = data$sex,residuals = residual_con)
p6 <-ggplot(data1, aes(x = Visit, y = residuals)) +geom_point(color ="#FFB6C1")+
  geom_boxplot(aes(group = Visit), alpha = 0.5,width = 0.5,color = "#A1C9F4", fill = "#A1C9F4") +
  labs(title = "Plots of normalized conditional residuals")    # conditional residuals
data1$sex <- factor(data1$sex, levels = c("male", "female"))
data1$Period <- factor(data1$Period, levels = c("A", "B"))
data1$Visit <- factor(as.character(data$Visit), levels = c("0", "1", "2", "3", "4","5","6","7"))
p7 <-ggplot(data1, aes(x = Visit, y = residuals)) +geom_boxplot(alpha = 0.5, color = "#A1C9F4", fill = "#A1C9F4", width = 0.5) +
  geom_point(color ="#FFB6C1") +facet_grid(. ~ sex) +  labs(title = "Plots of normalized conditional residuals per sex group") 
p8 <-ggplot(data1, aes(x = Visit, y = residuals)) + geom_boxplot(alpha = 0.5, color = "#A1C9F4", fill = "#A1C9F4", width = 0.5) +     geom_point(color ="#FFB6C1") +facet_grid(. ~ Period) +  labs(title = "Plots of normalized conditional residuals per Period") 
# p6  # p7 # p8
```

```{r}
#Histogram for EB estimates of random Intercepts
ranef_est <- ranef(model_new, which = "id", condVar = TRUE)
hist(ranef_est$"(Intercept)",main = "Histogram of the EB estimates of the random intercepts",freq = FALSE,xlab = "EB estimates (Intercept)",col = "darkgray",border = "#A1C9F4",breaks = 20)
#Histogram for EB estimates of random slopes
hist(ranef_est$"Visit",main = "Histogram of the EB estimates of the random slopes (Visit)",freq = FALSE,xlab = "EB estimates (Visit)",col = "darkgrey",border = "#A1C9F4",breaks = 20)
# Interpretation of the Random Effects Structure 
ranefs <- ranef(model_new)
plot(ranefs[, "(Intercept)"], ranefs[, "Visit"],xlab = "Random Intercepts", ylab = "Random Slopes",main = "Empirical Bayes Estimates")
qqnorm(model_new, ~ resid(., type = "n"), abline = c(0, 1),col = "#A1C9F4")    # QQ plot to check
acf(residuals(model_new, type = "response"))    # autocorrelation
```


# Disease Severity Analysis
## 1.Descriptive analysis
Q1.Present summary statistics for the sample at hand i.e., number of males/females, age distribution, etc.
```{r }
# Provides an overview of numeric and factor variables
summary(data)  
cat("\n### Visit Distribution\n")
table(data$Visit)  # Count of observations per visit time point
prop.table(table(data$Visit))  # Proportion per visit
cat("\n### Gender Distribution\n")
table(data$sex)  # Frequency of male (1) and female (2)
prop.table(table(data$sex))  # Proportion of males and females

cat("\n### Age Distribution\n")
range(data$Age)  # Min and max age
mean(data$Age)  # Average age
median(data$Age)  # Median age
sd(data$Age)  # Standard deviation

cat("\n### SNP Distribution\n")
table(data$SNP)  # Count of each SNP category (0,1,2)
prop.table(table(data$SNP))  # Proportion of each SNP category

cat("\n### Period Distribution\n")
table(data$Period)  # Count of patients in Period A vs. B
prop.table(table(data$Period))  # Proportion of each period

cat("\n### Disease Severity Distribution\n")
table(data$Severity)  # Count of low (0) vs high (1) severity
prop.table(table(data$Severity))  # Proportion of each severity level
```
data contains 2723 records
about 46% are male and 54% are female
participants of age 41 to 88 with an average of 64 years old and a Standard
deviation of 13.6 yrs
Proportion of each SNP category
50% SNP=0, 30% SNP=1, 20% SNP=2
Proportion of patients in each period :  nearly the same (50%-50%)
Proportion of patients for each severity level
76% severity=0 and 24% severity=1

For age, we categorized the continuous data into groups:
```{r }
# Create European Age Groups
data$Age_Group <- cut(data$Age, 
                      breaks = c(40, 45, 50, 55, 60, 65, 70, 75, 80, Inf), 
                      include.lowest = TRUE, 
                      right = FALSE, 
                      labels = c("40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"))

cat("\n### Age group Distribution\n")
table(data$Age_Group)  # Count of patients in Period A vs. B
prop.table(table(data$Age_Group))  # Proportion of each period

```

## 2.Exploratory analysis

### 2.1 cross table & graphing

Check the relationship between Severity and other variables. 
```{r }
cat("\n### Severity vs. Visit\n")
table(data$Severity.f, data$Visit)
prop.table(table(data$Severity.f, data$Visit), margin = 2)  # Proportion within Visit

cat("\n### Severity vs. Gender\n")
table(data$Severity.f, data$sex)  # Counts
prop.table(table(data$Severity.f, data$sex), margin = 2)  # Proportion within Gender

cat("\n### Severity vs. Age group\n")
table(data$Severity.f, data$Age_Group)  # Counts
prop.table(table(data$Severity.f, data$Age_Group), margin = 2)  # Proportion within Gender


cat("\n### Severity vs. SNP\n")
table(data$Severity.f, data$SNP)
prop.table(table(data$Severity.f, data$SNP), margin = 2)  # Proportion within SNP

cat("\n### Severity vs. Period\n")
table(data$Severity, data$Period)
prop.table(table(data$Severity, data$Period), margin = 2)  # Proportion within Period

```
### 2.2 Plots cross analysis 
Check the relationship between severity and other variables. 
```{r }
# Severity vs. Visit
ggplot(data, aes(x = as.factor(Visit), fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(x = "Visit", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
theme_minimal(base_size = 20) +  
  theme(axis.text.x = element_text(size = 16))  
# Severity vs. Gender
ggplot(data, aes(x = sex, fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Gender", x = "Gender", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  theme_minimal()

# Severity vs. Age group
ggplot(data, aes(x = Age_Group, fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Age Group", x = "Age Group", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), minimal = TRUE)

# Severity vs. Age (keep age as continuous variable and use Box Plot)
ggplot(data, aes(x = Severity.f, y = Age, fill = Severity.f)) +
  geom_boxplot(outlier.shape = 16, outlier.size = 2, alpha = 0.6) +
  labs(title = "Age Distribution by Severity", x = "Severity", y = "Age") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1")) +
  theme_minimal()

# Severity vs. SNP
ggplot(data, aes(x = as.factor(SNP), fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by SNP", x = "SNP", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  theme_minimal()

# Severity vs. Period
ggplot(data, aes(x = Period, fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Period", x = "Period", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  theme_minimal()
```

Observation summary

1. Visit: The plot shows a clear increasing trend in severity (proportion of 1s) over visits.
Initially (Visits 0–3), the severity proportion is low, but from Visit 4 onward, there is a sharp increase.
The trend suggests a potential nonlinear relationship between Visit and Severity.
A quadratic / cubic /   splines((piecewise,visit 0-3/4-7)  term for Visit might need to be added to improve model fit.

2. Gender also shows a notable effect, with females having a higher proportion of severe cases than males.
3. For age groups, the severity distribution appears relatively even, suggesting no strong relationship. 
4. SNP distribution is mostly even, but SNP type 2 has a slightly lower severity proportion. 
5. Period A shows a slightly higher severity proportion than Period B.


### 2.3 Interaction between Visit and other variables

Create facet plots where X-axis is always Visit, and facets are based on other categorical variables like Gender, Age Group, SNP, and Period.
```{r }

# Severity vs. Visit Faceted by Gender
ggplot(data, aes(x = as.factor(Visit), fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Visit (Faceted by Gender)", x = "Visit", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  facet_grid(. ~ sex) +  # Facet by Gender (one panel for each gender)
  theme_minimal()

# Faceted Box Plots: Age as a Continuous Variable
ggplot(data, aes(x = Severity.f, y = Age, fill = Severity.f)) +
  geom_boxplot(outlier.shape = 16, outlier.size = 2, alpha = 0.6) +
  labs(title = "Age Distribution by Severity (Faceted by Visit)", x = "Severity", y = "Age") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1")) +
  facet_grid(. ~ as.factor(Visit)) +  # Facet by Visit
  theme_minimal()


# Severity vs. Visit Faceted by Age Group
ggplot(data, aes(x = as.factor(Visit), fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Visit (Faceted by Age Group)", x = "Visit", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  facet_grid(. ~ Age_Group) +  # Facet by Age Group
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate X-axis labels


# Severity vs. Visit Faceted by SNP
ggplot(data, aes(x = as.factor(Visit), fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Visit (Faceted by SNP)", x = "Visit", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  facet_grid(. ~ SNP) +  # Facet by SNP (0,1,2)
  theme_minimal()


# Severity vs. Visit Faceted by Period
ggplot(data, aes(x = as.factor(Visit), fill = Severity.f)) +
  geom_bar(position = "fill") +
  labs(title = "Severity Distribution by Visit (Faceted by Period)", x = "Visit", y = "Proportion") +
  scale_fill_manual(values = c("#A1C9F4", "#FFB6C1"), name = "Severity") +
  facet_grid(. ~ Period) +  # Facet by Period (A vs. B)
  theme_minimal()

```

Observation summary

1. Gender-based differences is obvious in the severity progression over visits. Specifically, females experience sharper increase in severity proportion after the fourth visit, while males show a steadier trend. 
2. Age groups show no clear pattern, with severity levels evenly distributed across visits.
3. Weak interaction between SNP and visit. The effect of SNP is minor compared to the strong influence of visit on severity.
4. Moderate interaction between visit and period. Severity progression over visits differs between periods, with a steeper increase in earlier years (Period A). 


### 2.4 Individual difference / Random effect exploration

The plot clearly shows individual differences in predicted severity probabilities over visits. 
Patients are expected to have random intercept (baselines) and random slopes. 
Some patients progress quickly (steeper slopes), while others remain at lower severity probabilities for longer.

```{r }

# Fit a simple GLMM (no random slopes yet)
glmm_simple <- glmer(Severity.f ~ Visit + (1 | id), family = binomial, data = data)

# Get fitted probabilities
data$fitted_prob <- predict(glmm_simple, type = "response")

# Randomly select 20% of individuals :
set.seed(123)
selected_ids <- sample(unique(data$id), 100)
subset_data <- data[data$id %in% selected_ids, ]

# Plot fitted probabilities across visits for some patients
ggplot(subset_data, aes(x = Visit, y = fitted_prob, group = id, color = as.factor(id))) +
  geom_line(alpha = 0.7) +
  geom_point() +
  labs(title = "Individual Differences in Fitted Probabilities (100 Patients)",
       x = "Visit", y = "Predicted Probability of Severity") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend for clarity

```


## 3.Model Specification

  Elaborated GLMM Model:
  
  1. Mean Structure/Main effect: 
    - Response variable (Y): Severity level
    - Explanatory variables (X): Visit, gender, age, SNP, Period
    - Interaction: e.g. Visit * Sex
    - Nonlinear relationships quadratic or cubic

  2. Include random effects with different intercepts/slopes



## 4.Model Building & Selection

  Starting with an extremely elaborate model in GLMM, causes the software fail
  to converge mostly due to collinearity. 
  glmer() assumes that the random part completely captures the correlations,
  thus it uses a diagonal Var-Cov matrix just to estimate the variances 
  (diagonal elements).
  The model fails to converge if one considers quadratic terms for both
  'Visit' and 'Age'. Also, considering a polynomial of the third degree for
  'Visit' makes it hard to infer later in addition to some convergence problems.
  Therefore, We start with the following elaborate model. The nAGQ is chosen by 
  starting from 1 (Laplace approximation case) and increasing to reach stabilized
  fixed-effects parameters estimations, beside avoiding to become computationally
  intensive.

  the final model:
  Starting with an elaborate model, and considering all interactions and second
  order polynomials of the continuous variables, and removing the non-significant
  parameters by testing different combinations of these complex terms, finally
  the model 6 is the best one reached among all these models.


### Model 1: Severity.f ~ poly(Visit, 2) * sex + Age + SNP.f + Period + (Visit | id)

The elaborated model

```{r }

model.1 <- glmer(Severity.f ~ poly(Visit, 2) * sex + Age + SNP.f + Period + (Visit | id), 
                 data = data, family = "binomial", 
                 nAGQ = 1, control = glmerControl(optimizer = "bobyqa"))
summary(model.1)

```


### Model 2: Severity.f ~ Visit * sex + Age + SNP.f + Period + (Visit | id)
remove poly term
```{r }

model.2 <- glmer(Severity.f ~ Visit * sex + Age + SNP.f + Period + (Visit | id), 
                 data = data, family = "binomial", 
                 nAGQ = 1, control = glmerControl(optimizer = "bobyqa"))

summary(model.2)

```

### Model 1 vs. Model 2 --> model 2 wins, no need to add poly term
```{r }
anova(model.1, model.2)
```

### Model 6 (remove SNP&Age): Severity.f ~  Visit:Sex  + Period + (Visit | id) 
simplify the mean effect part: remove SNP & Age
```{r }

model.6 <- glmer(Severity.f ~ Visit * sex + Period + (Visit | id), 
                   data = data, family = "binomial", 
                   nAGQ = 1, control = glmerControl(optimizer = "bobyqa"))
summary(model.6)

```

### Model 2 vs. Model 6 --> model 6 wins
```{r }
anova(model.2, model.6)
```

### Model 68 (remove intercept) 
simplify the mean effect part: remove SNP & Age
```{r }

model.68 <- glmer(Severity.f ~ Visit * sex + Period + (1 | id), 
                   data = data, family = "binomial", 
                   nAGQ = 15, control = glmerControl(optimizer = "bobyqa"))
summary(model.68)

```

```{r }
anova(model.6, model.68)
```
## 5.Marginal Effects

### Model 60 (fit model 6 with `GLMMadaptive` ): Severity.f ~  Visit * sex + Period + (Visit | id)
Why model 60:
- GLMMadaptive uses Adaptive Gaussian-Hermite Quadrature (AGHQ) to approximate integrals over random effects, offering more flexibility than lme4::glmer()'s nAGQ.
- More accurate for small or skewed data: Handles complex random effect structures better.
- More stable variance estimates: glmer() with nAGQ = 1 may underestimate random effect variance.
- Cross-checking model stability: Comparing results from glmer() and GLMMadaptive helps assess GLMM reliability. 
If estimates are similar, glmer() is likely sufficient.

```{r }
# we have increased nAGQ to finally reach a stabilized model at nAGQ=11 
model.60 <- mixed_model(Severity.f ~ Visit * sex + Period, random = ~Visit|id, 
                        data = data, family =  "binomial", control = list(nAGQ = 11))


summary(model.60)

# The model is fitted using the adaptive Guassian quadrature approach which results
# to more reliable estimations.

```

### Parameters of Marginal Effects

Use GLMMadaptive to get mariginal effects: 

```{r }
# Marginalized parameters
marginal_coefs(model.60, std_errors = TRUE)
```

## 6.Model assumption check

The final model is model.6 

Investigate the distribution of the posterior modes of the random effects in the random intercepts model `model.6` . 

When fitting a random intercepts model (model.6), we typically assume that:
- Random intercepts and slopes follow a normal (Gaussian) distribution.
- There should be no extreme outliers or strong skewness in the distribution of random effects.

Investigating the posterior modes (also known as the empirical Bayes estimates) of these random effects helps us check if this assumption holds and whether there are extreme deviations.

### 6.1 dotplot

Conclusion: The model's normality assumption appears to be mostly valid:
- Random intercepts roughly follow a normal distribution, with no extreme outliers.
- Random slopes for Visit are tightly clustered near 0, indicating limited variability but still roughly symmetric.

```{r }
#Caterpillar plot for EB estimates of the random effects

library(lattice)
dotplot(ranef(model.6, which = "id", condVar = TRUE), 
        scales = list(y = list(alternating = 0))) # suppress id names

#The blue dots are the conditional modes with error bars. 
# the variations in the random slopes is much lesser than the variations in
# random intercepts. This means considering random slopes in the model is not as
# vital as considering the random intercepts

```

### 6.2 Histogram

The distribution of random intercepts roughly follows a normal distribution but is slightly right-skewed. 
The distribution of random slopes approximately follows a normal distribution.
```{r }
#Histogram for EB estimates of the random effects
# Extract random effects for the grouping factor "id"
re <- ranef(model.6, condVar = TRUE)$id  

# Convert to a data frame
random_effects_df <- as.data.frame(re)

# Check column names to verify the correct extraction
print(names(random_effects_df))

# Rename columns if necessary
colnames(random_effects_df) <- c("Random_Intercepts", "Random_Slopes")

# Set up 2-panel layout for histograms
par(mfrow = c(1, 2))  

# Histogram for random intercepts
hist(random_effects_df$Random_Intercepts, 
     main = "", 
     xlab = "Histogram of Random Intercepts", 
     freq = FALSE, 
     col = "gray", 
     breaks = 20)

# Histogram for random slopes
hist(random_effects_df$Random_Slopes, 
     main = "", 
     xlab = "Histogram of Random Slopes", 
     freq = FALSE, 
     col = "gray", 
     breaks = 20)

# Reset plot layout
par(mfrow = c(1, 1))


```

Conclusion: The normality assumption for random intercept and random slope is reasonable. Random intercepts are a little skewed to the right, while random slopes are slightly skewed to the left side.

## 7. Graphical visualization

Graphical visualization: Use visualization techniques to examine the mean structure


Compute fitted probabilities: 

### 7.1 Total without comparison in Sex
- The subject-specific fitted probabilities of high disease severity for ... 
```{r }
#Subject-specific fitted probabilities
data.fitted.1 <- data.frame(id = 8, 
                            Visit = 0:7, 
                            Age = 79,  
                            sex = "female",  
                            SNP = 0,
                            Period = "A"
                           )  
# predict probability
data.fitted.1$fit <- predict(model.6, newdata = data.fitted.1, type = "response")

xyplot(fit ~ Visit,
       data = data.fitted.1,  
       panel = function (...) {
           panel.xyplot(..., lty = c(1),
                        col = "black", lwd = 2, type = "l")
       }, 
       xlab = "Visits", 
       main = "Fitted conditional probability of high Disease Severity for individual",
       ylab = "Probability of high Disease Severity",
       scales = list(x = list(at = 0:7)))

```
- The fitted probabilities for the median individual...
```{r }
#Fitted probabilities for the median individual

data.fitted.2 <- data.frame(id = 10, 
                            Visit = 0:7, 
                            Age = 48.3,  
                            sex = "male",  
                            SNP = 1,
                            Period = "A"
                           ) 

# predict probability
data.fitted.2$fit <- predict(model.6, newdata = data.fitted.2, type = "response",re.form = NA)

xyplot(fit ~ Visit,
       data = data.fitted.2,  
       panel = function (...) {
           panel.xyplot(..., lty = c(1),
                        col = "black", lwd = 2, type = "l")
       }, 
       xlab = "Visits", 
       main = "Fitted conditional probability of high Disease Severity for median subject",
       ylab = "Probability of high Disease Severity",
       scales = list(x = list(at = 0:7)))

```

- The marginal (i.e., population averaged) fitted probabilities....

```{r }
log.odds <- predict(model.6, newdata = data, type = "link", re.form = NA)
meas.index <- match(data$id, unique(data$id))
B <- 100
probs.b <- matrix(NA, nrow = nrow(data), ncol = B)

for(bb in 1:B){
  b <- rnorm(length(unique(data$id)), 0, 
             sqrt(VarCorr(model.6)[[1]][1]))[meas.index]
  probs.b[, bb] <- plogis(log.odds + b) 
}

data$MargProbs <- data$fit <- rowMeans(probs.b)


xyplot(fit ~ Visit,
       data = data[data$id == 1, ],
       panel = function (...) {
           panel.xyplot(..., lty = c(1),
                        col = "black", lwd = 2, type = "l")
       }, 
       xlab = "Visits", 
       main = "Fitted conditional probability of high Disease Severity for marginal effect",
       ylab = "Probability of high Disease Severity",
       scales = list(x = list(at = 0:7)))
```


Individual subject vs. Median subject vs. Marginal 
```{r}
# Add a 'Group' column to distinguish the three sets of predictions
data.fitted.1$Group <- "Individual subject"
data.fitted.2$Group <- "Median subject"

# Compute marginal effect (averaging over random effects)
log.odds <- predict(model.6, newdata = data, type = "link", re.form = NA)
meas.index <- match(data$id, unique(data$id))
B <- 100
probs.b <- matrix(NA, nrow = nrow(data), ncol = B)

for (bb in 1:B) {
    b <- rnorm(length(unique(data$id)), 0, sqrt(VarCorr(model.6)[[1]][1]))[meas.index]
    probs.b[, bb] <- plogis(log.odds + b)
}

# Compute the average marginal probabilities
data$MargProbs <- rowMeans(probs.b)

# Extract subject id = 1 for marginal effect plotting and ensure all required columns exist
data.fitted.3 <- data[data$id == 1, c("id", "Visit", "MargProbs", "Age", "sex", "SNP", "Period")]

# Ensure column names match `data.fitted.1` and `data.fitted.2`
colnames(data.fitted.3)[colnames(data.fitted.3) == "MargProbs"] <- "fit"
data.fitted.3$Group <- "Marginal effect"

# Ensure 'Group' column is a factor with correct levels
data.fitted.3$Group <- factor(data.fitted.3$Group, levels = c("Individual subject", "Median subject", "Marginal effect"))

# Combine all three datasets
data.fitted <- rbind(data.fitted.1, data.fitted.2, data.fitted.3)

# Convert 'Group' into a factor for plotting
data.fitted$Group <- factor(data.fitted$Group, levels = c("Individual subject", "Median subject", "Marginal effect"))

# Plot the fitted probabilities
xyplot(fit ~ Visit, 
       groups = Group,  
       data = data.fitted,  
       type = "l",
       lty = c(1, 2, 3),  # Solid, dashed, dotted
       col = c("black", "black", "black"),  
       lwd = 2,  
       xlab = "Visits", 
       main = "Fitted Conditional Probability of High Disease Severity",
       ylab = "Probability of High Disease Severity",
       key = list(
           space = "top",  # Place legend above the lines
           columns = 3,    # Three labels in a single row
           lines = list(lty = c(1, 2, 3), col = c("black", "black", "black")),
           text = list(c("Individual subject", "Median subject", "Marginal effect"))
       ))

```

Summary:
Subject-Specific: Calculate and visualize the probability trend for a specific individual.  
Median Individual: Calculate the probability for an individual with only fixed effects(set random effect 0).
Marginal level: Calculate the population level average probability, integrating random effects and ignoring individual difference.


From the comparison above, we can see the clear individual differences. 
When we remove individual variability and consider only the median subject (only fixed effects), the predicted probability is the lowest level. 
The marginal effect falls in between, representing the population level average.


### 7.2 Comparison in Sex

Individual subject vs. Median subject vs. Marginal : Male vs. Female

Since we can see significant difference between male and female, we can visualize the probability difference:

- The subject-specific fitted probabilities of high disease severity for ... 
```{r }
#Subject-specific fitted probabilities
# Create dataset for ID = 8, predicting as both male and female
data.fitted.4 <- data.frame(
    id = 8, 
    Visit = 0:7, 
    Age = 79,  
    sex = c(rep("male", 8), rep("female", 8)),  # Predict for both male and female
    SNP = 1,
    Period = "A"
)

# Predict probability for both male and female
data.fitted.4$fit <- predict(model.6, newdata = data.fitted.4, type = "response")

# Convert 'sex' to factor for faceting
data.fitted.4$sex <- factor(data.fitted.4$sex, levels = c("male", "female"))

# Create a faceted plot for male vs. female
xyplot(fit ~ Visit | sex,
       data = data.fitted.4,  
       panel = function (...) {
           panel.xyplot(..., lty = c(1),
                        col = "black", lwd = 2, type = "l")
       }, 
       xlab = "Visits", 
       main = "Fitted Conditional Probability of High Disease Severity",
       ylab = "Probability of High Disease Severity",
       scales = list(x = list(at = 0:7)))


```
- The fitted probabilities for the median individual...
```{r }
#Fitted probabilities for the median individual

# Create dataset for ID = 10, predicting as both male and female
data.fitted.5 <- data.frame(
    id = 10, 
    Visit = 0:7, 
    Age = 48.3,  
    sex = c(rep("male", 8), rep("female", 8)),  # Predict for both male and female
    SNP = 1,
    Period = "A"
)

# Predict probability for both male and female
data.fitted.5$fit <- predict(model.6, newdata = data.fitted.5, type = "response", re.form = NA)

# Convert 'sex' to factor for faceting
data.fitted.5$sex <- factor(data.fitted.5$sex, levels = c("male", "female"))

# Create a faceted plot for male vs. female
xyplot(fit ~ Visit | sex,
       data = data.fitted.5,  
       panel = function (...) {
           panel.xyplot(..., lty = c(1),
                        col = "black", lwd = 2, type = "l")
       }, 
       xlab = "Visits", 
       main = "Fitted Conditional Probability of High Disease Severity for Median Subject",
       ylab = "Probability of High Disease Severity",
       scales = list(x = list(at = 0:7)))
```

- The marginal (i.e., population averaged) fitted probabilities.

```{r }
# Create dataset: Duplicate data for male and female predictions
data.marginal <- data
data.marginal <- data.marginal[rep(1:nrow(data.marginal), 2), ]  # Duplicate rows for both sexes
data.marginal$sex <- rep(c("male", "female"), each = nrow(data) )  # Assign male/female

# Predict log-odds for both male and female separately
log.odds <- predict(model.6, newdata = data.marginal, type = "link", re.form = NA)
meas.index <- match(data.marginal$id, unique(data.marginal$id))

# Number of bootstrap samples
B <- 100
probs.b <- matrix(NA, nrow = nrow(data.marginal), ncol = B)

# Simulate random effects and compute probabilities
for (bb in 1:B) {
    b <- rnorm(length(unique(data.marginal$id)), 0, sqrt(VarCorr(model.6)[[1]][1]))[meas.index]
    probs.b[, bb] <- plogis(log.odds + b)  # Convert log-odds to probabilities
}

# Compute the marginal probabilities as the mean over bootstrap samples
data.marginal$MargProbs <- rowMeans(probs.b)
data.marginal$fit <- data.marginal$MargProbs  # Align with previous plotting structure

# Convert 'sex' to factor for faceting
data.marginal$sex <- factor(data.marginal$sex, levels = c("male", "female"))

# Plot the marginal probabilities by gender
xyplot(fit ~ Visit | sex,
       data = data.marginal[data.marginal$id == 1, ],  # Select one subject (id = 2)
       panel = function (...) {
           panel.xyplot(..., lty = c(1),
                        col = "black", lwd = 2, type = "l")
       }, 
       xlab = "Visits", 
       main = "Marginal Fitted Probability of High Disease Severity by Gender",
       ylab = "Probability of High Disease Severity",
       scales = list(x = list(at = 0:7)))


```

Individual subject vs. Median subject vs. Marginal 
```{r}
# Duplicate dataset for male and female
data.marginal <- data
data.marginal <- data.marginal[rep(1:nrow(data.marginal), 2), ]  # Duplicate rows for both sexes
data.marginal$sex <- rep(c("male", "female"), each = nrow(data))  # Assign male/female

# Predict log-odds for both male and female separately
log.odds <- predict(model.6, newdata = data.marginal, type = "link", re.form = NA)
meas.index <- match(data.marginal$id, unique(data.marginal$id))

# Number of bootstrap samples
B <- 100
probs.b <- matrix(NA, nrow = nrow(data.marginal), ncol = B)

# Simulate random effects and compute probabilities
for (bb in 1:B) {
    b <- rnorm(length(unique(data.marginal$id)), 0, sqrt(VarCorr(model.6)[[1]][1]))[meas.index]
    probs.b[, bb] <- plogis(log.odds + b)  # Convert log-odds to probabilities
}

# Compute the marginal probabilities as the mean over bootstrap samples
data.marginal$MargProbs <- rowMeans(probs.b)
data.marginal$fit <- data.marginal$MargProbs  # Align with previous plotting structure
data.marginal$Group <- "Marginal effect"

# ---- Create datasets for Individual and Median predictions ---- #
# Individual subject
data.fitted.4$Group <- "Individual subject"
data.fitted.5$Group <- "Median subject"

# Extract subject id = 1 for marginal effect plotting and ensure all required columns exist
data.fitted.6 <- data.marginal[data.marginal$id == 1, c("id", "Visit", "MargProbs", "Age", "sex", "SNP", "Period", "Group")]
colnames(data.fitted.6)[colnames(data.fitted.6) == "MargProbs"] <- "fit"

# Combine all three datasets
data.fitted_sex <- rbind(data.fitted.4, data.fitted.5, data.fitted.6)

# Ensure 'Group' column is a factor with correct levels
data.fitted_sex$Group <- factor(data.fitted_sex$Group, levels = c("Individual subject", "Median subject", "Marginal effect"))

# Convert 'sex' to a factor
data.fitted_sex$sex <- factor(data.fitted_sex$sex, levels = c("male", "female"))

# ---- Plot all three probability types while faceting by gender ---- #
xyplot(fit ~ Visit | sex,
       groups = Group,  
       data = data.fitted_sex,  
       type = "l",
       lty = c(1, 2, 3),  # Solid, dashed, dotted
       col = c("black", "black", "black"),  
       lwd = 2,  
       xlab = "Visits", 
       main = "Fitted Conditional Probability of High Disease Severity by Gender",
       ylab = "Probability of High Disease Severity",
       key = list(
           space = "top",  # Place legend above the lines
           columns = 3,    # Three labels in a single row
           lines = list(lty = c(1, 2, 3), col = c("black", "black", "black")),
           text = list(c("Individual subject", "Median subject", "Marginal effect"))
       ))


```


## 8.Conclusion: Summarize key findings based on the final validated model.
Report the parameter estimates with the corresponding 95\% CIs.

model.6 recap: 
```{r }
model.6 <- glmer(Severity.f ~ Visit * sex + Period + (Visit | id), 
                   data = data, family = "binomial", 
                   nAGQ = 1, control = glmerControl(optimizer = "bobyqa"))
summary(model.6)

```

95% confidence intervals (CIs) for the log-odds
```{r }
# Table of estimates with 95% CI
se <- sqrt(diag(vcov(model.6)))
(tab <- cbind(Est = fixef(model.6), 
              LL = fixef(model.6) - 1.96 * se, 
              UL = fixef(model.6) + 1.96 * se))
 
```

95% confidence intervals (CIs) for odds ratios
```{r }
tab_OR <- exp(tab)  # Convert log-odds to odds ratios (exp)
tab_OR
```
Marginal effect
```{r }
# we have increased nAGQ to finally reach a stabilized model at nAGQ=11 
model.60 <- mixed_model(Severity.f ~ Visit * sex + Period, random = ~Visit|id, 
                        data = data, family =  "binomial", control = list(nAGQ = 11))
summary(model.60)

# The model is fitted using the adaptive Guassian quadrature approach which results
# to more reliable estimations.
```

Use GLMMadaptive to get mariginal effects: 

```{r }
# Marginalized parameters
marginal_coefs = marginal_coefs(model.60, std_errors = TRUE)
marginal_coefs
```

```{r }
# 1. Get coefficient matrix from marginal_coefs()
x <- marginal_coefs(model.60, std_errors = TRUE)
tab <- as.data.frame(x$coef_table)

# 2. Add rownames as a column (optional, for clarity)
tab$Term <- rownames(tab)

# 3. Calculate OR and CI
tab$OR <- exp(tab$Estimate)
tab$Lower_CI <- exp(tab$Estimate - 1.96 * tab$Std.Err)
tab$Upper_CI <- exp(tab$Estimate + 1.96 * tab$Std.Err)

# 4. Reorder columns for clean output
tab_clean <- tab[, c("Term", "Estimate", "Std.Err", "z-value", "p-value", "OR", "Lower_CI", "Upper_CI")]

# 5. Optional: round for display
tab_clean <- transform(tab_clean,
  Estimate = round(Estimate, 4),
  Std.Err = round(Std.Err, 4),
  `z-value` = round(`z-value`, 2),
  OR = round(OR, 4),
  Lower_CI = round(Lower_CI, 4),
  Upper_CI = round(Upper_CI, 4)
)

tab_clean

```
